Architectures for workflow integration in science gateways

### 2016-04-15 paper.tex
Updates in section 3.6

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software components and interactions. Software components include
science gateway and infrastructure in all architectures, and workflow
service, workflow pool and agent are involved in some
architectures. The workflow engine itself is represented by a specific
symbol. Software interactions among these components are described in
the next Section. Abstract interactions are specific types of
interactions that may be implemented by various different software
interactions. Red color is used to represent the components and
interactions that need to be specifically developed to integrate
workflow engines. Tasks refer to computing tasks that are created by
the workflow engine and executed on the infrastructure. Data
represents any type of file, or database that is involved in the
workflow execution and stored on the infrastructure while the workflow
executes. It does not cover workflow parameters such as strings,
numbers, etc. The reminder of this section describes the different
types of interactions involved in the architectures, and details each
architecture.

@@

See Figure~\ref{archi:nested}. A nested workflow is an abstract
architectural pattern where an activity of a master workflow is itself
a workflow that is executed by another engine. Nested workflows can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model as this is
the most used architecture. The master workflow is also called
meta-workflow. The nested workflow might be described with a different
language than the one used to describe the master workflow.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They were
also used by the SHIWA Science Gateway to implement so-called
Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution. It
corresponds to the nested worfklow with service invocation in
Figure~\ref{archi:nested}. The parent workflow engine is WS-PGRADE,
invoked as a service in the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\ref{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through a ``submission service'' (step
\texttt{2}, interaction \texttt{d}). Each of these engines can submit
jobs and transfer data to a distributed computing infrastructure (DCI,
step \texttt{3}, interactions \texttt{b} and \texttt{c$_2$}).

Nested workflows
are also often used implicitly to wrap scripts as black-boxes while
these scripts actually correspond to complete workflows.

\todo{instantiation is for foreign workflows, i.e. 2 services. We could have another one with native workflows too.}

\todo{both workflow engines have to access the infrastructure. If they
  run on 2 infrastructures, c1 has to be duplicated.}

\todo{define parent and child engine}

\todo{explain that such an abstract architecture has to be concretized to allow for comparison with other architectures. Say also that we could envisage other concretizations but we don't include them as we don't have any experience with these systems.}

@@

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. \todo{update colors and find a more relevant way to compute a grand total without favoring criteria that have more metrics than the others (e.g. normalize each criterion between 0 and 1). Same comment for each sub-total. M1 might be the same as integration.}}

>>>>

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures, and workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol. Software interactions among these
components are described in the next Section. \emph{Abstract}
interactions are specific types of interactions that may be
implemented by various different software interactions. An
architecture that has an abstract interaction is an abstract
architecture. Red color is used to represent the components and
interactions that need to be specifically developed to integrate
workflow engines. Tasks refer to computing tasks that are created by
the workflow engine and executed on the infrastructure. Data
represents any type of file, or database that is involved in the
workflow execution and stored on the infrastructure while the workflow
executes. It does not cover workflow parameters such as strings,
numbers, etc. The reminder of this section describes the different
types of interactions involved in the architectures, and details each
architecture.

@@

See Figure~\ref{archi:nested}. In nested workflows (Figure~\ref{archi:nested}-Left), an activity of a
\emph{parent} workflow executed by the \emph{parent} engine is itself
a \emph{child} workflow that is executed by a \emph{child} engine. The
parent and child engines might use different workflow description
languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through
interaction \texttt{*$_1$}. The science gateway also communicate with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions as well,
\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows in the science
gateway through interaction \texttt{a$_2$}, and installs software on
infrastructures through interactions \texttt{a$_1$}.

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model(see
Figure~\ref{archi:nested}-Right) as this is the most used
architecture. In such an instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfer user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow activities as any other command-line tool. Nested
workflows were also notably used by the SHIWA Science Gateway to
implement so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\ref{archi:nested}. The
parent workflow engine is WS-PGRADE, invoked as a service in the
Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\ref{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (DCI, step \texttt{3}, interactions \texttt{b} and
\texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

@@

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. \todo{update colors and find a more relevant way to compute a grand total without favoring criteria that have more metrics than the others (e.g. normalize each criterion between 0 and 1). Same comment for each sub-total. M1 might be the same as integration. Update nested workflow as infrastructures were separated.}}
